|

### 2024-08-21 reproducibility/manuscript/manuscript.qmd
docs(manuscript): add preprocessing methods
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

@@

We used the default options for projecting the vector fields from scVelo models.
Unlike scVelo, Pyro-Velocity uses statistics derived from posterior samples of
the denoised spliced gene expression and posterior samples of the velocity
estimation for building the cell state transition matrix estimates using cosine
similarity. Pyro-Velocity uses the same projection method as scVelo for
projecting the transition matrix into the two-dimensional vector field on the
user-provided embedding space.

>>>>

## Single-cell data preprocessing {#sec-methods-preprocessing}

We used scanpy and scVelo to handle the data input and output; thus, both h5ad
and loom files generated by velocyto and kallisto [@Melsted2021-ap] are
supported. The fully mature PBMC dataset was processed with the same procedure
proposed in a review paper [@Bergen2021-qz]
(<https://scvelo.readthedocs.io/perspectives/Perspectives/>). We reproduced this
procedure using the scVelo package and raw read counts of the same top three
dynamical genes NKG7, IGHM, and GNLY with the best likelihoods. The pancreas
dataset was processed with scVelo using the following options

```python
scv.pp.filter_and_normalize(
  adata=adata,
  min_shared_counts=30, 
  n_top_genes=2000,
)
scv.pp.moments(
  adata,
  n_pcs=30,
  n_neighbors=30,
)
```

The same top variable genes with raw spliced and unspliced read counts were used
as input for the Pyro&thinsp;-Velocity model. The original LARRY dataset of in
vitro Hematopoiesis containing $130,887$ cells was first filtered to remove
cells without LARRY barcoding. $49,302$ cells were recovered after this step
with at least one LARRY barcode. For simplicity, we termed this filtered dataset
with multiple cell fate (multi-fate) as the full dataset. Based on this dataset,
we created two datasets with uni-fate progression toward monocyte or neutrophil
based on the lineage LARRY barcodes and time information. Namely, we selected
sets of cells with a single LARRY barcode, spanning three time points (day 2, 4,
6), and all the cells from the last time point (day 6) belong to a unique cell
type (either monocyte or neutrophil). The two uni-fate datasets were combined to
represent the bi-fate LARRY dataset. The multi-fate full dataset was processed
using the same options as the pancreas dataset; the rest of the uni-fate and
bi-fate datasets were processed using the following parameters

```python
scv.pp.filter_and_normalize(
  adata=adata,
  n_top_genes=2000,
  min_shared_counts=20,
)
scv.pp.moments(adata)
```

## scVelo model {#sec-methods-scvelo}

We benchmarked the dynamical RNA velocity model implemented in scVelo `v0.2.4`
for the pancreas and the four LARRY datasets using the same user options

```python
scvelo.tl.recover_dynamics(
  data=adata, 
  n_jobs=30,
)
scvelo.tl.velocity(
  data=adata, 
  mode="dynamical",
)
```

Then, we tested a set of user options, including the neighboring cell numbers
and the top variable gene numbers, in the pancreas dataset to explore the
stability of the scVelo dynamical model. For the fully mature PBMC dataset, we
followed the notebook proposed by the original authors
<https://scvelo.readthedocs.io/perspectives/Perspectives/>, i.e., we used the
stochastic RNA velocity model implemented in scVelo with the top three
likelihood genes. The latent time from scVelo was computed using their provided
function

```python
scvelo.tl.latent_time(
  data=adata,
)
```

@@

We used the default options for projecting the vector fields from scVelo models.
Unlike scVelo, Pyro&thinsp;-Velocity uses statistics derived from posterior samples of
the denoised spliced gene expression and posterior samples of the velocity
estimation for building the cell state transition matrix estimates using cosine
similarity. Pyro&thinsp;-Velocity uses the same projection method as scVelo for
projecting the transition matrix into the two-dimensional vector field on the
user-provided embedding space.
